Dynamic Prediction with fPCA 

### 2022-10-31 FDAgroup.Rmd
updated FDA group presentation slides

---
title: "Dynamic Prediction in Functional Concurrent Regression "
output: beamer_presentation
date: "2022-10-27"
---

@@

@@

$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}$$
where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.

@@

- Use FPCA to estimate $f_0$, $\Gamma$ and $\sigma_{\epsilon}$

@@

$$\hat{\boldsymbol{Y}} = \boldsymbol{B}^T\hat{\boldsymbol{u}} $$

@@

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.3] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.3, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.3]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.3", 
       x = "time", y = "")

@@

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.5] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.5, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.5]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.5", 
       x = "time", y = "")

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.8] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.8, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.8]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.8", 
       x = "time", y = "")

>>>>

---
title: "Dynamic Prediction with fPCA "
output: beamer_presentation
date: "2022-11-3"
---

@@

- Goal: to predict future track based on observed track

@@

$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}, \hspace{0.5cm} j = 1...J_i$$
where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.

@@

- When there is no covariate in the model, this is essentially a fPCA problem.

  - $\boldsymbol{B}$ is a matrix of eigenfunctions
  - $\boldsymbol{u}$ is a matrix of PC scores/loadings

- Use fPCA to estimate $f_0$, $\Gamma$ and $\sigma_{\epsilon}$

@@

$$\hat{\boldsymbol{Y}} = \hat{\boldsymbol{f}}_0+ \boldsymbol{B}^T\hat{\boldsymbol{u}} $$

@@

```{r, pred_func}
# write a function for prediction up to t
# df_test: datafram with columns argvals, subj, y, observation up to tm

DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){
  Ntest = length(unique(df_test$subj))
  # prediction
  pred1 <- predict(fpca_fit, newdata = df_test)
  # Eigen functions
  Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ]
  Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores)
  # every column is a new function
  Yhat <- data.frame(Yhat)
  colnames(Yhat) <- pred1$rand_eff$subj
  df_pred <- Yhat %>% 
    mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>% 
    pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>%
    mutate(type = "pred")

  return(df_pred)
}
```

```{r t0_3}
df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.3))

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.3", 
       x = "time", y = "")
```

```{r}
df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.5), tm = 0.5)

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.5", 
       x = "time", y = "")

@@

```{r}
df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.8), tm = 0.8)

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.8", 
       x = "time", y = "")
